---
title: "Multivariate statistics"
subtitle: "Examination"
author: "Samuel Orso"
date: "`r Sys.Date()`"
output: 
 prettydoc::html_pretty:
  theme: architect
  highlight: github
  toc: true
  df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exam instruction
Please note the following instructions for your examination:    

1. The exam is closed book (no books, no slides, ...), but a cheat sheet of 6 pages (3 sheets A4) is allowed.   
2. Computers and cell phones, or any other communications system, are forbidden.   
3. A calculator of your choice is allowed.   
4. The exam lasts 120 minutes.   
5. You must answer with a blue or black pen (no other colors, no pencil).    
6. The exam is written in both french and english and you can decide to answer in french or in english (but only one).   
7. The exam is seperated into multiple quetions (QCM) and open questions:   

    * QCM: There are 15 questions with 5 possible answers each. There is only **one** correct answer. One correct answer is 1 point, one wrong answer is -0.2 point, no answer is 0 point. Two or more answers is one wrong answer.
    * Open questions: You must answer on the exam sheets within the dedicated space, external sheets are not considered during the correction. It is a supreme necessity that you justify each of your answers.   
    
# Examples
### QCM

**Question:**   

What is the signification of the level $\alpha$ when testing hypothesis?   
_Que signifie le niveau $\alpha$ dans un test d'hypothèse?_

1. The probability of making correct inference.  
  _La probabilité de faire une inférence correcte._   
2. The risk of failing to reject $H_0$ when $H_0$ is false.   
  _Le risque de ne pas rejeter $H_0$ lorsque $H_0$ est fausse._   
3. The probability of accepting $H_1$.   
  _La probabilité d'accepter $H_1$._  
4. The risk of rejecting $H_0$ when $H_0$ is true.   
  _Le risque de rejeter $H_0$ lorsque $H_0$ est vraie._
5. The risk of rejecting $H_0$ when $H_0$ is false.   
  _Le risque de rejeter $H_0$ lorsque $H_0$ est fausse._  
  
**Answer:** (not provided with the exam)   

When testing an hypothesis, the level $\alpha$ is the risk of rejecting the null-hypothesis when it is true, also known as the Type I error. Clearly answer 4. is correct and answers 1.,2. and 5. are wrong. Answer 3. is also wrong because: first, you never accept an hypothesis but fail to reject it, second, the statement is unconditionnal of the true nature of $H_0$ (you are not saying whether $H_0$ is true or false).


### Open questions

**Question:**   
You receive this `R` output:
```{r, echo=F, message = F, warning = F}
require(car)
fit_lm <- lm(income ~ age + sex + population, data = Chile)
```

```{r}
summary(fit_lm)
```

Based on this output, write down the estimated regression model and specify which regression coefficient are significant at a level of $\alpha=5\%$.

_Vous recevez cet ouput de `R`. Ecrivez le modèle de régression estimé et spécifiez quel(s) coefficient(s) de régression sont significatifs à un niveau $\alpha=5\%$._

**Answer:** (not provided with the exam)   

The estimated regression model is given by:
$$ \hat{y} = 22'490 - 100.7\;\text{age} + 4'455\;\text{sex}_{\text{M}} + 0.0836\;\text{population}, $$
where $\hat{y}$ are the estimated `income` based on this regression model. All the regression coefficients are significant at a level of $\alpha=5\%$.


**Question:**   
Suppose that `sex` is a factor with two levels: `F` for female and `M` for male. Based on your previous answer, what is the marginal effect of being a male on the income with this regression model?   

_Supposez que `sex` est un facteur avec deux niveaux: `F` pour femme et `M` pour homme. En vous basant sur votre réponse précédente, quelle est l'effet marginal d'être un homme sur les revenus dans ce modèle de régression?_

**Answer:** (not provided with the exam)  

Marginally being a male increases of 4'455 units the income (we do not know in what units the income is expressed) based on this regression model.

